k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2023-05-04 paper.Rmd
more editing

When defining a new variable $y = (y_1, \ldots, y_N)$ as the squared deviation of each observation from the mean $\overline{\mathbf{x}}$, i.e. $y_i = (x_i - \overline{\mathbf{x}})^2$ ($i = 1, \dots, N$), the sample variance of $x$ can be reformulated as the mean of $y$:

@@

Research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict with each other (Ferligoj & Batagelj, 1992). As $\mathit{SSE_{kplus}}$  is a novel criterion, it is so far unknown whether maximizing it can yield a partitioning where both constituting criteria are similar between groups, i.e., whether these objectives are aligned or in conflict. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ (i.e., $w_1$ = $w_2$ = 1 in the computation of $\mathit{SSE_{kplus}}$), which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree. This result indicates that the objectives $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ are not in conflict with each other. Moreover, in Figure 1 there is a high concentration of partitions in the upper-right corner where the two objectives are in high agreement.

@@

Here, $w_t$ is the weight associated with the $t$'th moment and $\overline{\mathbf{X}}^{(k)}_{T \times P} = \{\overline{x}_{tj}^{(k)}\}_{T \times P}$ is a centroid matrix containing the mean values for all k-plus variables in the $k$'th group, i.e.

@@

Thus, when setting the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ has the same form as the standard k-means criterion $\mathit{SSE}$, but it is computed on different data. Computing the k-plus criterion is then done by first creating the additional k-plus variables, appending them as new columns to the original data matrix, and computing the $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

@@

In general, the weighting scheme for k-plus anticlustering can become highly complex as it might encompass an arbitrary number of distribution moments, and hence, an arbitrary number of weights. While an exhaustive evaluation of the influence of weights is beyond the scope of the current paper, a small scale simulation study was conducted to investigate the influence of weights for a selected number of moments. The target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minimizing differences with regard to group means, variances, skewness and kurtosis. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented k-plus variables---of the input data $\mathbf{X'}_{N \times \mathit{TP}}$ before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardized matrix $\mathbf{X'}_{N \times \mathit{TP}}$ was used. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favoring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favoring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

To compare the different weighting schemes, `r length(results)` data sets from were generated from a normal distribution ($M = 0$, $\mathit{SD} = 1$). The number of variables $P$ was determined randomly for each data set and varied between 1 and 5; the number of anticlusters $K$ was also determined randomly for each data set and varied between 2 and 4; the sample size $N$ randomly varied between 20 and 100 with the restriction that $N$ had to be divisible by $K$ so that equal-sized groups could be created. For each data set, $\mathit{SSE_{kplusGeneralized}}$ was optimized using the local maximum algorithm, once for each of the six competing weighting schemes.

@@

In Simulation 2, the performance of two k-plus criteria was compared to the performance of two traditional anticlustering criteria. "Standard" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplus}}$, thus minimizing differences with regard to group means and variances. "Generalized" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplusGeneralized}}$ using $T = 4$, thus minimizing differences with regard to group means, variances, skewness and kurtosis. The unweighted standardization approach, that performed best in Simulation 1, was used as the benchmark implementation for k-plus anticlustering that was compared to the competing objectives. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, and also (iii) a random allocation of elements to groups.

@@

As a secondary purpose of Simulation 2, I investigated the degree to which partitions identified by the different anticlustering methods showed similarity. Specifically, I investigated whether different anticlustering solutions showed any overlap regarding the placement of individual elements in groups. In cluster analysis, high quality partitions according to different objectives tend to overlap to a large degree, but it is an open question whether this result also holds for different anticlustering partitions. To tackle this issue, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each data set and for each $K$, i.e., 30,000 times.

@@

Table 3 illustrates the degree of overlap between anticlustering partitions that were identified in Simulation 2. Overall, it is shown that there was hardly any similarity between anticlustering partitions according to the different criteria. The average ARI was almost 0 in all cases---indicating no similarity of partitions whatsoever. Moreover, the ARI between two anticlustering partitions hardly differed from the ARI between an anticlustering partition and a random partition. This constitutes an interesting and novel result, illustrating that high between-group similarity can be achieved via very different partitionings.

@@

Anticlustering is a partitioning method that aims for high between-group similarity and high within-group heterogeneity. In this paper, I presented the k-plus criterion for anticlustering, which focuses on maximizing between-group similarity. K-plus represents between-group similarity as discrepancies in distribution moments between groups: means, variances, skewness, kurtosis, and higher order moments. Depending on the application, a different subset of these objectives can be included in the anticlustering process. The k-plus criterion is an extension of the classical k-means criterion, which only reflects how similar groups are with regard to their means. Interestingly, however, the k-plus extension does not change the original k-means criterion because additional criteria can be incorporated by augmenting the input data.

@@

The present paper focused on the new k-plus criterion for anticlustering, but only briefly considered the multicriterion nature of this criterion. A simple unweighted sum approach was used to optimize the k-plus objective, while more sophisticated algorithmic procedures exist to tackle multicriterion optimization problems [e.g., @brusco2009cross]. The sole focus on the objective function followed two reasonings. First, the applications and the simulations showed that the unweighted standardization approach yielded satisfying results that probably suffice in most applications. Moreover, it is possible that k-plus anticlustering has a fortunate solution space where many suitable partitions exist that fulfill multiple criteria to a satisfactory degree (as suggested by Figure 1). Still, more sophisticated approaches may yield theoretically improved results, as well as the opportunity to deliberately select among partitionings that may be preferred according to different k-plus objectives. @brusco2019 gave an example of how direct algorithms can be employed to optimize multiple criteria in anticlustering applications aiming for high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar algorithmic treatment. A second reason for focusing on the k-plus objective per se---independently from the algorithm that is used to optimize it---was to obtain a fair comparison to existing anticlustering objectives. In the simulation study, differences in performance could not have been interpreted if the objective function as well as the algorithm used to optimize it had differed between anticlustering methods. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering.

As an additional outlook, it should be noted that because the covariance is defined as an expected value, the k-plus logic can be extended to minimize differences with regard to covariances, a possibility that was not investigated in the current paper. In this case, an additional variable has to be computed for each pair of variables $x = (x_1, \dots, x_N)$ and $y = (y_1, \dots, y_N)$, computed as $(x_i - \overline{\mathbf{x}}) (y_i -\overline{\mathbf{y}})$. Future research should examine how the k-plus approach can be adapted to including covariances. The unweighted standardization approach presented here may serve as a starting point for investigation.

@@

Anticlustering is usually implemented by reversing an objective function used in classical cluster analysis. Whereas cluster analysis seeks homogeneity within groups and separation between groups, the reversal results in heterogeneity within groups and similarity between groups. While tightly linked in a mathematical sense, an important semantic difference is that cluster analysis is concerned with finding "true" structure in the data, e.g., by defining a taxonomy on naturally occurring groups [@everitt1979]. Anticlustering instead imposes structure on the data, while the anticlusters themselves do not represent any underlying "true" pattern. Correspondingly, different---but similarly adequate---anticlustering partitions tend to look very different, as was also shown in Simulation 2. In contrast, different clustering solutions tend to look similar because a good clustering is characterized by assigning similar elements into the same cluster and dissimilar elements into different clusters. A good anticlustering solution however does not depend on assigning individual items either to the same or a different group, independent of their pairwise dissimilarity.[^dispersion] This distinction can also be clarified using Figure 1, where we observe a high concentration of suitable anticlustering partitions but only few suitable clustering solutions. An astonishing `r mean(df$kmeans / max(df$kmeans) >.999) * 100`% of all partitions are within 0.1% of the global maximum in $\mathit{SSE}$, i.e., the best k-means anticlustering solution. In contrast, there are only `r sum(min(df$kmeans)/ df$kmeans > .90)` partitions (`r mean(min(df$kmeans)/ df$kmeans > .90)*100`%) that approximate the global minimum in $\mathit{SSE}$ (i.e., the best k-means clustering solution) by even 10%. Thus, even though the problems are based on the same objective function, cluster analysis and anticlustering exhibit a stark discrepancy with regard to the solution space of partitions. Interestingly, the k-plus objective introduces an additional distinction between cluster analysis and anticlustering. Reversing the k-plus anticlustering objective no longer yields a useful clustering interpretation, because identifying groups that are different with regard to higher order moments does not constitute a clustering. Thus, even though k-plus extends the classical k-means clustering criterion, it is a pure anticlustering objective.

>>>>

When defining a new variable $y = (y_1, \ldots, y_N)$ as the squared deviation of each observation from the mean $\overline{\mathbf{x}}$, i.e. $y_i = (x_i - \overline{\mathbf{x}})^2$, the sample variance of $x$ can be reformulated as the mean of $y$:

@@

Research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict (Ferligoj & Batagelj, 1992). As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown whether maximizing it can yield a partitioning where both constituting criteria are similar between groups, i.e., whether these objectives are aligned or in conflict. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ (i.e., $w_1$ = $w_2$ = 1 in the computation of $\mathit{SSE_{kplus}}$), which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

@@

Here, $w_t$ is the weight associated with the $t$'th moment and $\overline{\mathbf{X}}^{(k)}_{T \times P} = \{\overline{x}_{tj}^{(k)}\}_{T \times P}$ is a matrix containing the mean values for all k-plus variables in the $k$'th group, i.e.

@@

Thus, when setting the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ has the same form as the standard k-means criterion $\mathit{SSE}$, but is based on different data. Here, computing the k-plus criterion is done by first creating the additional k-plus variables, appending them as new columns to the original data matrix, and then computing the $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

@@

In general, the weighting scheme for k-plus anticlustering can become highly complex as it might encompass an arbitrary number of distribution moments, and hence, an arbitrary number of weights. While an exhaustive evaluation of the influence of weights is beyond the scope of the current paper, a small scale simulation study was conducted to investigate the influence of weights for a selected number of moments. The target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minimizing differences with regard to group means, variances, skewness and kurtosis. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented k-plus variables---of the input data before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardized data matrix was used. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favoring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favoring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

To compare the different weighting schemes, `r length(results)` data sets from were generated from a normal distribution ($M = 0$, $\mathit{SD} = 1$). The number of variables $P$ was determined randomly for each data set and varied between 1 and 5; the number of anticlusters $K$ was also determined randomly for each data set and varied between 2 and 4; the sample size $N$ randomly varied between 20 and 100 with the restriction that $N$ had to be divisible by $K$. For each data set, $\mathit{SSE_{kplusGeneralized}}$ was optimized using the local maximum algorithm, once for each of the six competing weighting schemes.

@@

In Simulation 2, the performance of two k-plus criteria was compared to the performance of two traditional anticlustering criteria. "Standard" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplus}}$, thus minimizing differences with regard to group means and variances. "Generalized" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplusGeneralized}}$ using $T = 4$, thus minimizing differences with regard to group means, variances, skewness and kurtosis. The unweighted standardization approach that performed best in Simulation 1 was used as the benchmark implementation for k-plus anticlustering that was compared to the competing objectives. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, and also (iii) a random allocation of elements to groups.

@@

As a secondary purpose of Simulation 2, I investigated the degree to which partitions identified by the different anticlustering methods showed similarity. Specifically, I investigated whether different anticlustering solutions showed any overlap regarding the placement of individual elements in groups. In cluster analysis, high quality partitions according to different objectives tend to overlap to a large degree, but it is an open question whether this result also holds for different anticlustering partitions. To tackle this issue, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each data set and for each $K$.

@@

Table 3 illustrates the degree of overlap between anticlustering partitions that were identified in Simulation 2. Overall, it is shown that there was hardly any similarity between anticlustering partitions according to the different criteria. The average ARI was almost 0 in all cases---indicating no similarity of partitions whatsoever. Moreover, the ARI between two anticlustering partitions hardly differed from the ARI between an anticlustering partition and a random partition. This constitutes an interesting result, illustrating that high between-group similarity can be achieved via very different partitionings.

@@

Anticlustering is a partitioning method that aims for high between-group similarity and high within-group heterogeneity. In this paper, I presented the k-plus criterion for anticlustering, which focuses on maximizing between-group similarity. K-plus represents between-group similarity as discrepancies in distribution moments between groups: means, variances, skewness, kurtosis, and higher order moments. Depending on the application, a different subset of these objectives can be included in the anticlustering process. The k-plus criterion is an extension of the classical k-means criterion, which only reflects how similar groups are with regard to their means. Interestingly, when using the unweighted sum approach presented here, the k-plus extension does not change the original k-means criterion, because additional criteria can be incorporated by augmenting the input data.

@@

The present paper focused on the new k-plus criterion for anticlustering, but only briefly considered the multicriterion nature of this criterion. A simple unweighted sum approach was used to optimize the k-plus objective, while more sophisticated algorithmic procedures exist to tackle multicriterion optimization problems [e.g., @brusco2009cross]. The sole focus on the objective function followed two reasonings. First, the applications and the simulations showed that the unweighted standardization approach yielded satisfying results that probably suffice in most applications. Moreover, it is possible that k-plus anticlustering has a fortunate solution space where many suitable partitions exist that fulfill multiple criteria to a satisfactory degree (as suggested by Figure 1). Still, more sophisticated approaches may yield theoretically improved results, as well as the opportunity to deliberately select among partitionings that may be preferred according to different k-plus objectives. @brusco2019 gave an example of how direct algorithms can be employed to optimize multiple criteria in anticlustering applications aiming for high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar algorithmic treatment. Specifically, it should be investigated if the bicriterion approach by Brusco et al. can be transformed into a multicriterion algorithm for the k-plus objective. A second reason for focusing on the k-plus objective per se---independently from the algorithm that is used to optimize it---was to obtain a fair comparison to existing anticlustering objectives. In the simulation study, differences in performance could not have been interpreted if the objective function as well as the algorithm used to optimize it had differed between anticlustering methods. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering. Future research should investigate whether this advantage also holds for other data distributions than normal. For example, it may be hypothesized that a generalized k-plus criterion is even more useful when applied to data that is not symmetric or contains outliers, because it is able to specifically incorporate the skewness and kurtosis as criteria in the anticlustering process.

As an additional outlook, it should be noted that because the covariance is defined as an expected value, the k-plus logic can be extended to minimize differences with regard to covariances, a possibility that was not investigated in the current paper. In this case, we have to compute an additional variable for each of pair of variables whose covariance should be similar between groups. In future research, the unweighted standardization approach presented here may serve as a starting point for investigating covariance similarity in anticlustering applications.

@@

Anticlustering is usually implemented by reversing an objective function used in classical cluster analysis. Whereas cluster analysis seeks homogeneity within groups and separation between groups, the reversal results in heterogeneity within groups and similarity between groups. While tightly linked in a mathematical sense, an important semantic difference is that cluster analysis is concerned with finding "true" structure in the data, e.g., by defining a taxonomy on naturally occurring groups [@everitt1979]. Anticlustering instead imposes structure on the data, while the anticlusters themselves do not represent any underlying "true" pattern. Correspondingly, different---but similarly adequate---anticlustering partitions tend to look very different, as was also shown in Simulation 2. In contrast, different clustering solutions tend to look similar because a good clustering is always characterized by assigning similar elements into the same cluster and dissimilar elements into different clusters. A good anticlustering solution however does not depend on assigning individual items either to the same or a different group, independent of their pairwise dissimilarity.[^dispersion] This result was observed in Simulation 2 and can also be clarified using Figure 1. Here, we observe a high concentration of suitable anticlustering partitions but only few suitable clustering solutions. An astonishing `r mean(df$kmeans / max(df$kmeans) >.999) * 100`% of all partitions are within 0.1% of the global maximum in $\mathit{SSE}$, i.e., the best k-means anticlustering solution. In contrast, there are only `r sum(min(df$kmeans)/ df$kmeans > .90)` partitions (`r mean(min(df$kmeans)/ df$kmeans > .90)*100`%) that approximate the global minimum in $\mathit{SSE}$ (i.e., the best k-means clustering solution) by even 10%. Thus, even though the problems are based on the same objective function, cluster analysis and anticlustering exhibit a stark discrepancy with regard to the solution space of partitions. Interestingly, the k-plus objective introduces an additional distinction between cluster analysis and anticlustering. Reversing the k-plus anticlustering objective no longer yields a useful clustering interpretation, because identifying groups that are different with regard to higher order moments does not constitute a clustering. Thus, even though k-plus extends the classical k-means clustering criterion, it is a pure anticlustering objective.
